Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-03 application.tex
rough version 2 done

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed
by assuming vertical derivatives can be neglected in
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
% we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
where $\hat{\eta} = \eta/H^2\Omega_k$ is the 
dimensionless diffusion coefficient.

@@

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
% where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
% $\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
% and Eq. \ref{eta_def}. 
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the 
% In order to utilize our linear results, which assumes $t_c$ is a
% constant, we need to make further 
% approximations. Specifically, we consider vertically-isothermal disks
% and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
% H$, where $\Sigma$ is the surface density at the fiducial
% radius of interest. For example, the dimensionless diffusion
% coefficient becomes 
% \begin{align}\label{diff_coeff_dimensionless}
%   \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
%     c_v\Omega_k}. 
% \end{align}
% Furthermore, we will choose a temperature-dependent opacity model for
% simplicity, so that $\kappa_d=\kappa_d(T)$.
dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{1}{\hat{\eta}}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that the thermal relaxation time depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane. %The latter is due to the density-dependence in the diffusion time. 
%is then a constant for vertically-isothermal disks, 
%and we can directly apply previous results.

@@

We therefore take Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see that growth timescales are $\sim
30$---40 orbits, and the VSI operates from a few AU to tens of
AU. Importantly, it is necessary to consider smaller scales as $r$
decreases as the disk generally becomes optically
thick.

Fig. \ref{mmsn_overall} is roughly consistent with the 
mid-plane thermal relaxation timescales shown in
Fig. \ref{mmsn_bcrit_bcool}. For example, for $\khat=10$ we find 
$\beta \gtrsim \beta_\mathrm{crit}$ for $r\lesssim 10$AU at $z=0$; and
Fig. \ref{mmsn_overall} show stabilization inside $\sim 8$AU. % The mode
% operates interior to the expected minimum radius,

% probably because the disk bcomes optically thin away from the
% midplane.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the fundamental VSI mode in
    the MMSN.  
    \label{mmsn_overall}}    
\end{figure}

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ only for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with 
significantly reduced growth rates (when scaled by the local
rotation).  
%: Fig. \ref{mmsn_overall}) show that
%the growth time at

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer than the case with $\beta < \beta_\mathrm{crit}$ for
all $z$: we find $t_\mathrm{grow}\sim 40P_\mathrm{orb}$ at $r=50$AU,
but $t_\mathrm{grow}\sim 400P_\mathrm{orb}$ at $r=5$AU.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_5AU}
  \caption{Vertical velocity perturbations of the VSI in the MMSN at 
    $r=50$AU (top) and $r=5$AU (bottom). The real (imaginary) part of 
    $\delta v_z$ is shown as
    the solid (dotted) curves. The eigenfunction has been normalized
    such that $\imag(\delta v_z)=0$ at the vertical boundaries, and
    $\mathrm{max}|\delta v_z|=1$. \label{mmsn_eigenvz}}
\end{figure}

>>>>

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed
by assuming vertical derivatives can be neglected in
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
% we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
% where $\hat{\eta} = \eta/H^2\Omega_k$ is the 
% dimensionless diffusion coefficient.

@@

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
% where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
% $\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
% and Eq. \ref{eta_def}. 
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the 
% In order to utilize our linear results, which assumes $t_c$ is a
% constant, we need to make further 
% approximations. Specifically, we consider vertically-isothermal disks
% and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
% H$, where $\Sigma$ is the surface density at the fiducial
% radius of interest. For example, the dimensionless diffusion
% coefficient becomes 
% \begin{align}\label{diff_coeff_dimensionless}
%   \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
%     c_v\Omega_k}. 
% \end{align}
% Furthermore, we will choose a temperature-dependent opacity model for
% simplicity, so that $\kappa_d=\kappa_d(T)$.
dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\Omega_k}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that the thermal relaxation time depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane. %The latter is due to the density-dependence in the diffusion time. 
%is then a constant for vertically-isothermal disks, 
%and we can directly apply previous results.

@@

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see that growth timescales are $\sim
30$---40 orbits, and the VSI operates from a few AU to tens of
AU. Importantly, it is necessary to consider smaller scales as $r$
decreases as the disk generally becomes optically thick.

Fig. \ref{mmsn_overall} is roughly consistent with the 
mid-plane thermal relaxation timescales shown in
Fig. \ref{mmsn_bcrit_bcool}. For example, for $\khat=10$ we find 
$\beta \gtrsim \beta_\mathrm{crit}$ for $r\lesssim 10$AU at $z=0$; and
Fig. \ref{mmsn_overall} show stabilization inside $\sim 8$AU.

%The mode % operates interior to the expected minimum radius, 
% probably because the disk bcomes optically thin away from the
% midplane.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the fundamental VSI mode in
    the MMSN. These are calculated using the scale and height
    dependent thermal relaxation given by Eq. \ref{beta_mmsn_simp}. 
    \label{mmsn_overall}}    
\end{figure}

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with 
significantly reduced growth rates (when scaled by the local
rotation).  
%: Fig. \ref{mmsn_overall}) show that
%the growth time at

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer than the case with $\beta < \beta_\mathrm{crit}$ for
all $z$: we find $t_\mathrm{grow}\sim 40$ orbits at $r=50$AU,
but $t_\mathrm{grow}\sim 400$ orbits at $r=5$AU. Thus, introducing an
optically thick midplane reduces the efficiency of the VSI despite
$\beta < \beta_\mathrm{crit}$ away from the midplane.

% The grow
% mode because it involves vertical motion across $z=0$, which is

@@

% Introducing an optically thick mid-plane may not completely surpress
% the VSI because the vertical entropy gradient, which usually provides
% stabilization, vanishes at $z=0$. (Note that $N_z^2\propto z^2$ while
% vertical shear $d\Omega^2/dz \propto z$)

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_5AU}
  \caption{Vertical velocity perturbations of the VSI in the MMSN at 
    $r=50$AU (top) and $r=5$AU (bottom). These are calculated using the height
    dependent thermal relaxation given by Eq. \ref{beta_mmsn_simp}. The real (imaginary) part of 
    $\delta v_z$ is shown as
    the solid (dotted) curves. The eigenfunction has been normalized
    such that $\imag(\delta v_z)=0$ at the vertical boundaries, and
    $\mathrm{max}|\delta v_z|=1$. \label{mmsn_eigenvz}}
\end{figure}

### 2015-03-03 linear_iso.tex
rough version 2 done

The VGSB is a self-consistent framework but extrapolating results 
obtained from the VGSB to the global disk has limitations. This is
because within the VGSB framework, $d\Omega/dz$ can be freely
specified. In effect, by adopting the VGSB framework, we are studying
dynamics in the standard shearing box subject to an additional
azimuthal force $\delta v_zr d \Omega/dz$ with $d\Omega/dz$ as an
input parameter. For consistency with the global disk, we choose
$d\Omega/dz$ in the VGSB to be that due to the thermal structure in
the global disk (Eq. \ref{vertical_shear}). However, 
% This is different to the global problem because in
% that case, the vertical shear is a result of the 
the background radial disk structure is ignored elsewhere in the VGSB.

We further discuss caveats of the VGSB in Appendix \ref{global_corr}.
However, we find that for perturbations with small radial lengthscales
($k_xH\gg 1$) and fast thermal relaxation ($\beta\ll1$), which
characterizes the VSI, the VGSB framework is adequate for the purpose
of a simple analytical discussion.

>>>>

The VGSB is a self-consistent framework but extrapolating results 
obtained from the VGSB to the global disk has limitations. This is
because within the VGSB framework, $d\Omega/dz$ can be freely
specified. In effect, by adopting the VGSB framework, we are studying
dynamics in the standard shearing box subject to an additional
azimuthal force $\delta v_zr d \Omega/dz$ with $d\Omega/dz$ as an
input parameter. For consistency with the global disk, we choose
$d\Omega/dz$ in the VGSB to be that due to the thermal structure in
the global disk (Eq. \ref{vertical_shear}). However, 
% This is different to the global problem because in
% that case, the vertical shear is a result of the 
the background radial disk structure is ignored elsewhere in the VGSB
(which is done by setting $\hat{g}_c=0$).

We further discuss caveats of the VGSB in Appendix \ref{global_corr}.
We find that for perturbations with small radial lengthscales
($k_xH\gg 1$) and fast thermal relaxation ($\beta\ll1$), which
characterizes the VSI, the VGSB framework is adequate for the purpose
of a simple analytical discussion.

### 2015-03-03 numerical.tex
rough version 2 done

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
% disk model is a nearly-vertically
% isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
% $\epsilon=0.1$. 
We first calculate the VSI in a neutrally
stratified disk ($N_z^2\equiv0$) with rapid thermal relaxation by
setting $\gamma=\Gamma=1.011$ and $\beta=10^{-3}$. Other disk
parameters are $(p,q,\epsilon)=(-1.5,-1,0.1)$ and the vertical domain
is $\zmax = 5H$. The disk is nearly vertically isothermal with
$c_s(\zmax)\simeq 0.94 c_s(0)$. Fig. \ref{omega_z} shows the
equilibrium rotation profile away from the disk mid-plane.

@@

{\bf ASK. Growth rates are expected to diminish for both $\khat\to 0$
  and $\khat\to \infty$. This is because for $\khat\ll 1$ the radial
  wavelenth is large, but such disturbances are 
  stabilized by by epicyclic motions ($|\omega|\sim
  \kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
  there is no energy change when fluid elements are perturbed at fixed
  cylindrical radius whilst conserving its angular momentum (since we
  consider axisymmetric perturbations).  
}

>>>>

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
% disk model is a nearly-vertically
% isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
% $\epsilon=0.1$. 
We first calculate the VSI in a neutrally
stratified disk ($N_z^2\equiv0$) with rapid thermal relaxation by
setting $\gamma=\Gamma=1.011$ and $\beta=10^{-3}$. Other disk
parameters are $(p,q,\epsilon)=(-1.5,-1,0.1)$ and the vertical domain
is $\zmax = 5H$. The disk is nearly vertically isothermal with
$c_s(\zmax)\simeq 0.94 c_s(0)$. Fig. \ref{omega_z} shows that the 
equilibrium rotation decreases slightly away  from the disk mid-plane.

@@

{\bf Growth rates are expected to diminish for both $\khat\to 0$
  and $\khat\to \infty$. This is because for $\khat\ll 1$ the radial
  wavelenth is large, but such disturbances are 
  stabilized by by epicyclic motions ($|\omega|\sim
  \kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
  there is no energy change when fluid elements are perturbed at fixed
  cylindrical radius whilst conserving its angular momentum (since we
  consider axisymmetric perturbations).  
}

### 2015-03-03 paper.tex
rough version 2 done

From these studies it is clear that the VSI is generally sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually assumed for typical disk
models,  
% disk thermodynamics except
% in the special case of neutrally stratified disks. 
it is 
important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the viability of
the VSI in realistic disks. % Numerical simulations of the VSI are
% challenging, however, because global disk simulations should be
% carried out while resolving radial scales much smaller than the local
% disk scale height.    
%On the other hand, 
Should a given disk be unstable to the VSI, it must pass through the
linear instability first. Thus, studying the thermodynamic dependence
of the linear VSI is essential. % Furthermore, linear
% stability calculations are essential to complement future numerical simulations.

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of early
analysis of \cite{urpin98} and \cite{urpin03} which accounts for thermal
relaxation, but considered vertically localized disturbances. Our
vertically global approach is necessary to study disturbances with
vertical wavelengths comparable to or larger than the disk scale
height. Such disturbances were  found by \cite{nelson13} and
\cite{stoll14} to dominate their numerical simulations.  We thus focus
on vertically global perturbations.

@@

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially-local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic thermal timescale below which
the VSI is expected to operate. In \S\ref{numerical} we numerically
solve the linearized fluid equations in the radially-local
approximation, but relax all other assumptions made in our analytical approach.   
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebulae (MMSN) and find that the VSI may operate in the outer
parts of the MMSN. We summarize and discuss our results in
\S\ref{summary}, including caveats and possible directions for future
work.

>>>>

From these studies it is clear that the VSI is generally sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually assumed for typical disk
models,  
% disk thermodynamics except
% in the special case of neutrally stratified disks. 
it is 
important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the viability of
the VSI in realistic disks. % Numerical simulations of the VSI are
% challenging, however, because global disk simulations should be
% carried out while resolving radial scales much smaller than the local
% disk scale height.    
%On the other hand, 
Should a given disk be unstable to the VSI, it must pass through the
linear instability first. Thus, understanding the thermodynamic dependence
of the linear VSI is essential. % Furthermore, linear
% stability calculations are essential to complement future numerical simulations.

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of early
analysis of \cite{urpin98} and \cite{urpin03}, which accounts for thermal
relaxation, but considered vertically localized disturbances. Our
vertically global approach is necessary to study disturbances with
vertical wavelengths comparable to or larger than the disk scale
height. Such disturbances were  found by \cite{nelson13} and
\cite{stoll14} to dominate their numerical simulations.  We thus focus
on vertically global perturbations which, in fact, we find to be the
dominant mode in protoplanetary disks with a realistic estimate for
the thermodynamic timescale.

@@

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially-local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic thermal timescale below which
the VSI is expected to operate. In \S\ref{numerical} we numerically
solve the linearized fluid equations in the radially-local
approximation, but relax all other assumptions made in our analytical approach.   
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebulae (MMSN) by modeling thermal relaxation
timescales based on realistic disk properties, and find that the VSI may
operate in the outer parts of the MMSN. We summarize and discuss our
results in \S\ref{summary}, including caveats and possible directions
for future work.
